### 2016-07-06 introduction.tex
Typos etc.

A major problem with weak isolation is that its semantics in the
context of user programs is not well-understood. The original
proposal~\cite{gray1976} defines multiple ``degrees'' of weak
isolation in terms of implementation details such as the nature and
duration of locks held in each case. The ANSI SQL 92 standard defines
four levels of isolation (including serializability) in terms of
various undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is
required to prevent. While this is an improvement, it requires
programmers to be prescient about the possible ways various
undesirable phenomena might manifest in their applications, and
determine, in each case, if the phenomenon can be allowed without
violating application invariants. This is understandably hard,
especially in the absence of any formal underpinning to define weak
isolation semantics.  Adya~\cite{adyaphd} presents the first formal
definitions of some well-known isolation levels in the context of a
sequentially consistent (SC) database.  However, there has been little
progress relating Adya's system model to a formal operational
semantics or a proof system that can facilitate rigorous correctness
arguments. Consequently, reasoning about weak isolation remains an
error prone endeavour, with major database vendors~\cite{postgresiso,
  mysqliso, oracleiso} continuing to document their isolation levels
primarily in terms of the undesirable phenomena a particular isolation
level may induce, placing the burden on the programmer to determine
application correctness.

>>>>

A major problem with weak isolation is that its semantics in the
context of user programs is not well-understood. The original
proposal~\cite{gray1976} defines multiple ``degrees'' of weak
isolation in terms of implementation details such as the nature and
duration of locks held in each case. The ANSI SQL 92 standard defines
four levels of isolation (including serializability) in terms of
various undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is
required to prevent. While this is an improvement, it requires
programmers to be prescient about the possible ways various
undesirable phenomena might manifest in their applications, and in
each case determine if the phenomenon can be allowed without violating
application invariants. This is understandably hard, especially in the
absence of any formal underpinning to define weak isolation semantics.
Adya~\cite{adyaphd} presents the first formal definitions of some
well-known isolation levels in the context of a sequentially
consistent (SC) database.  However, there has been little progress
relating Adya's system model to a formal operational semantics or a
proof system that can facilitate rigorous correctness arguments.
Consequently, reasoning about weak isolation remains an error prone
endeavour, with major database vendors~\cite{postgresiso, mysqliso,
oracleiso} continuing to document their isolation levels primarily in
terms of the undesirable phenomena a particular isolation level may
induce, placing the burden on the programmer to determine application
correctness.

### 2016-07-06 motivation.tex
Typos etc.

The specification of an isolation level encodes its interference
characteristics as constraints over the $\hbZ$ relation. However, for
this to be useful in reasoning about programs, the rely-guarantee
framework should be able to use the specification to determine if an
interference is valid or not, allowing the programmer to only
focus on the former. Our second observation is that this is possible
if the reasoning framework adequately tracks $\hbZ$ at each program
point, while preserving $\hbZ$ constraints as invariants between
program points. An interference that leads to the violation of $\hbZ$
constraints (i.e., an invalid interference) is thus automatically
prohibited. For instance, consider the program point after the write
to \C{B} in \C{Wd1}. The expected invariant ($\phi$) at that program
point is shown below ($\committed$ stands for ``committed''):

>>>>

The specification of an isolation level encodes its interference
characteristics as constraints over the $\hbZ$ relation. However, for
this to be useful in reasoning about programs, the rely-guarantee
framework should be able to use the specification to determine if an
interference is valid or not, allowing the programmer to only focus on
valid interferences. Our second observation is that this is possible
if the reasoning framework adequately tracks $\hbZ$ at each program
point, while preserving $\hbZ$ constraints as invariants between
program points. An interference that leads to the violation of $\hbZ$
constraints (i.e., an invalid interference) is thus automatically
prohibited. For instance, consider the program point after the write
to \C{B} in \C{Wd1}. The expected invariant ($\phi$) at that program
point is shown below ($\committed$ stands for ``committed''):

### 2016-07-06 opsem.tex
Typos etc.

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we will use to formalize the intuitions
presented in the previous section. Natural numbers, (shared) variables
and arithmetic expressions constitute the syntactic class of
expressions ($e$).  Commands ($c$) include $\cskip$, assignment
statements, transaction (\C{txn}) lexical blocks, and their sequential
and parallel composition. We let $T_i$ for $i \in \mathbb{N}$ range
over transaction identifiers. When it is evident we are referring to a
transaction, we use the number $i$ instead of $T_i$ for identification
(\eg, in $\C{txn}\langle i \rangle$). For notational convenience, we
let $t$ range over both expressions and commands.

@@

Fundamental to our development is the notion of a trace invariant
($\I$). $\I$ is a function from traces ($\E$) to first-order logical
formula ($\Prop$) that define well-formedness
constraints over traces.  The machine takes a step only if the
resulting trace satisfies the constraints imposed by $\I$. This
behaviour is captured by the auxiliary reduction rule
\rulelabel{E-Aux} that factors out the trace extension aspect of the
evaluation by abstracting away the operation-specific behaviour as a
function that generates an appropriate effect. We let $\mathcal{F}$
denote this function.  \rulelabel{E-Aux} uses $\mathcal{F}$ to
generate a new effect and extend the trace ($\E = (\A,\visZ)$)
\emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
(i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
execution that runs to completion, every small-step preserves the
well-formedness of a trace, thus ensuring the invariance of $\I$.
Note that the semantics makes no assumptions about $\I$ other than its
type. As such, it can be instantiated with any trace-parametric
proposition that expresses constraints over the given trace. For
instance, consider the $\psi_{RC}$ specification from
\S\ref{sec:motivation}, but with bounded $T_1$ and $T_2$ instantiated
with \C{Wd1} and \C{Wd2}, respectively. The instantiated specification
is the following term:\vspace*{-8pt}

@@

A more notable source of non-determinism, however, is the
\rulelabel{E-Aux} rule, which allows the machine to expose an
arbitrary subset ($S$) of existing effects ($\A$) to the incoming
operation. In other words, the machine is not obligated to reveal the
effects of all previous operations to an incoming operation. This
relaxation allows the abstract machine to model the semantics of
weakly-consistent data stores. For instance, operations issued to an
eventually consistent ({\sc ec}) replicated store could be dispatched
to different replicas whose states may not be in any well-defined
relationship. By allowing operations to witness arbitrary subsets of
the global state, the semantics models the weak visibility properties
of such stores; we elaborate on the implication of this style of
definition in \S\ref{sec:store-consistency}.  Stronger visibility
properties can be expressed by imposing well-formedness constraints
over $\visZ$ via the trace invariant ($\I$). Since the abstract
machine is obligated to satisfy $\I$ at every step of the execution,
operations are guaranteed to experience the level of isolation
specified by $\I$.  Thus, in executions that run to completion, the
abstract machine models a store that provides the required levels of
isolation.  Notably, the machine achieves this without defining an
operational semantics for isolation levels, instead solely relying on
their declarative characterization as trace well-formedness
constraints to enforce isolation guarantees.
\S\ref{sec:ansi-isolation} shows the specification of various ANSI SQL
isolation levels stated as trace well-formedness constraints.

>>>>

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we will use to formalize the intuitions
presented in the previous section. Natural numbers, (shared) variables
and arithmetic expressions constitute the syntactic class of
expressions ($e$).  Commands ($c$) include $\cskip$, assignment
statements, transaction (\C{txn}) lexical blocks, and their sequential
and parallel composition. We let $T_i$ for $i \in \mathbb{N}$ range
over transaction identifiers. When it is evident we are referring to a
transaction, we use the number $i$ instead of $T_i$ for identification
(\eg in $\C{txn}\langle i \rangle$). For notational convenience, we
let $t$ range over both expressions and commands.

@@

Fundamental to our development is the notion of a trace invariant
($\I$). $\I$ is a function from traces ($\E$) to first-order logical
formulas ($\Prop$) that define well-formedness
constraints over traces.  The machine takes a step only if the
resulting trace satisfies the constraints imposed by $\I$. This
behaviour is captured by the auxiliary reduction rule
\rulelabel{E-Aux} that factors out the trace extension aspect of the
evaluation by abstracting away the operation-specific behaviour as a
function that generates an appropriate effect. We let $\mathcal{F}$
denote this function.  \rulelabel{E-Aux} uses $\mathcal{F}$ to
generate a new effect and extend the trace ($\E = (\A,\visZ)$)
\emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
(i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
execution that runs to completion, every small-step preserves the
well-formedness of a trace, thus ensuring the invariance of $\I$.
Note that the semantics makes no assumptions about $\I$ other than its
type. As such, it can be instantiated with any trace-parametric
proposition that expresses constraints over the given trace. For
instance, consider the $\psi_{RC}$ specification from
\S\ref{sec:motivation}, but with bounded $T_1$ and $T_2$ instantiated
with \C{Wd1} and \C{Wd2}, respectively. The instantiated specification
is the following term:\vspace*{-8pt}

@@

A more notable source of non-determinism, however, is the
\rulelabel{E-Aux} rule, which allows the machine to expose an
arbitrary subset ($S$) of existing effects ($\A$) to the incoming
operation. In other words, the machine is not obligated to reveal the
effects of all previous operations to an incoming operation. This
relaxation allows the abstract machine to model the semantics of
weakly-consistent data stores. For instance, operations issued to an
eventually consistent ({\sc ec}) replicated store could be dispatched
to different replicas whose states may not be in any well-defined
relationship. By allowing operations to witness arbitrary subsets of
the global state, the semantics models the weak visibility properties
of such stores; we elaborate on the implication of this style of
definition in \S\ref{sec:store-consistency}.  Stronger visibility
properties can be expressed by imposing well-formedness constraints
over $\visZ$ via the trace invariant ($\I$). Since the abstract
machine is obligated to satisfy $\I$ at every step of the execution,
operations are guaranteed to experience the level of isolation
specified by $\I$.  Thus, in executions that run to completion, the
abstract machine models a store that provides the required levels of
isolation.  Notably, the machine achieves this without defining an
operational semantics for isolation levels, instead solely relying on
their declarative characterization as trace well-formedness
constraints to enforce isolation guarantees.
\S\ref{sec:ansi-isolation} specifies various ANSI SQL isolation levels
stated as trace well-formedness constraints.
